---
title: "Palmer_Penguins_Assignment_2023"
output: html_document
date: "2023-11-22"
---

# **Palmer Penguins Assignment 2023**

**Question 2** (Q1a) Write a data analysis pipeline in your .rmd RMarkdown file. You should be aiming to write a clear explanation of the steps as well as clear code

*Introduction*

The data set contains size measurements for three species of penguins found on three islands in the Palmer Archipelago, Antarctica (Gorman et al 2014). This data pipeline aims to explore the Palmer Penguin data set further; particularly investigating the interspecific variation in flipper length (i.e. looking to see if there is a relationship between flipper length and species).

The data must first be cleaned appropriately.

```{r}
# View the structure of the dataset
str(penguins)

# Check for missing values in the original dataset
summary(is.na(penguins))

# Remove NAs from 'species' and 'flipper_length_mm', and ensure flipper length is always positive (i.e. checking data integrity)
cleaned_penguins <- penguins[complete.cases(penguins$species, penguins$flipper_length_mm) & penguins$flipper_length_mm >= 0, ]

# Check the cleaned dataset
summary(is.na(cleaned_penguins))
```

Secondly, the data must be explored. This can be achieved by creating an exploratory figure

```{r}

# Jitter plot (exploratory figure) for flipper length of each species
ggplot(data = cleaned_penguins, aes(x = species, y = flipper_length_mm)) +
  geom_jitter(aes(color = species),
              width = 0.2, 
              alpha = 0.7) +
  labs(title = "Jitter Plot Showing the Distribution of Flipper Lengths by Species",
       x="Species",
       y="Flipper Length (mm)")+
  scale_color_manual(values = c("darkorange","darkgreen","blue3"))

```

*Hypothesis:*

H0: There is no difference in mean flipper length between species

H1: At least one of the species's mean flipper length significantly differs from at least one other species's mean flipper length

Based on the exploratory figure, I predict that the flipper length of Gentoo penguins will significantly differ from both Chinstrap and Adelie. Therefore, I predict that there is a difference in mean flipper length between species.

*Statistical Methods:*

As the investigation is between a categorical explanatory variable (i.e. species) and continuous response variable (i.e. flipper length), an Analysis of Variance (ANOVA) will be used. The ANOVA is used to test whether individuals chosen from different groups, on average, are more different than individuals chosen from the same group. In this case, investigating whether individuals from different species are, on average, more different regarding flipper length than individuals from the same species.

```{r}
#Fitting a linear model to the cleaned data
flipper_mod1 <- lm(flipper_length_mm ~ species, cleaned_penguins)
summary(flipper_mod1)
```

```{r}
#Running the ANOVA function
anova(flipper_mod1)
```

Based on this ANOVA, a post-hoc comparison can be conducted to identify which species are different from each other. The post-hoc comparison is a Tukey-Kramer test (also known as a Tukey Honest Significance Test). It is also called a Tukey HSD for short.

```{r}
#Run the TukeyHSD 
TukeyHSD(aov(flipper_mod1))
```

*Results:*

Box plots can be used to visualise the results of the ANOVA. This is because the box plot demonstrates the distribution of each group (species) and shows and potential differences in central tendency and spread.

```{r}
#Box plot for flipper length and species
ggplot(data = cleaned_penguins, aes(x = species, y = flipper_length_mm, colour = species,  alpha = 0.7)) +
  geom_boxplot(fill = "white", color = "black") +  # Customize boxplot appearance
  geom_point(position = position_jitter(), size = 2) +  # Customize point appearance
  scale_color_manual(values = c("darkorange", "darkgreen", "blue3")) +  # Set custom colors
  labs(title = "Distribution of Flipper Lengths by Species",
       x = "Penguin Species", y = "Flipper Length (mm)") +
  theme_minimal()  

```

An alternative visualisation method is the violin plot. Violin plots can also be used to visualise the results of the ANOVA as they show the distribution of the data along with the summary statistics and are useful for comparing different groups, which in this case is species.

```{r}
#Violin plot for flipper length and species
ggplot(data = cleaned_penguins, aes(x = species, y = flipper_length_mm, fill = species, alpha = 0.7)) +
  geom_violin(color = "black", scale = "width", trim = FALSE) +  # Use geom_violin for violin plot
  geom_point(position = position_jitter(width = 0.2), size = 2) +  # Customize point appearance
  scale_fill_manual(values = c("darkorange", "darkgreen", "blue3")) +  # Set custom colors
  labs(title = "Distribution of Flipper Lengths by Species",
       x = "Penguin Species", y = "Flipper Length (mm)") +
  theme_minimal()
```

*Discussion:*
The coefficient table demonstrate that the mean flipper length for: Adelie penguins is 189.954 mm, Chinstrap penguins is 195.824 mm, and Gentoo penguins is 217.187 mm. Ecologically, Gentoo penguins are deep divers whereas Adelie and Chinstrap penguins are shallow divers. This behavioural segregation appears to be anatomically reflected; with Gentoo penguinis having longer flipper lengths to support them in deeper swimming (Trivelpiece et al 1987).
The adjusted R-squred value of 0.777 indicates that 78% of the variation in the dataset is explained by differences between species. This is high for biological data.

The ANOVA table shows a p value of less than 0.05. This shows that there is evidence to reject the null hypothesis because our p-value is less than 0.05 (conventional alpha level). This result is interpreted as: at least one of the species of penguin differs from at least one other species of penguin.

To identify which species are different from each other, a Tukey HSD was conducted. The results show that for each pairwise comparison the p-value is less than 0.05. This means that the null hypothesis is rejected (i.e. two group means are the same). Therefore, each species differs in flipper length significant from every other species (i.e. all three species differ from each other significantly).

These results make biological sense as different species have different anatomical/physical/physiological adaptations. As such, each species of penguin is expected to have different flipper size as they may be adapted to different ecological niches.

*Conclusion:*
Based on the results and analysis, we can conclude that flipper length varies between species significantly. This intraspecific variation is biologically coherent as ecological discrepencies between species lead to anatomical adaptations and differences. In order of shortest to longest flipper length: Adelie, Chinstrap, and Gentoo. 
